---
title: "cBioPortalData: Quick Start Guide"
author: "Marcel Ramos & Levi Waldron"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{cBioPortal Quick-start Guide}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
---
# Installation

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(cBioPortalData)
library(AnVIL)
library(rapiclient)
```

## GitHub Packages

Changes in the development branch of `rapiclient` are yet to be merged to the
CRAN package. Please use the devel version of `rapiclient` along with the
`AnVIL` Bioconductor package.

```{r, eval=FALSE}
library(BiocManager)
install(c("Bioconductor/AnVIL_rapiclient", "Bioconductor/AnVIL"))
install("waldronlab/cBioPortalData")
```

```{r, eval=FALSE}
library(cBioPortalData)
library(AnVIL)
library(rapiclient)
```

# API representation

Obtaining the cBioPortal API representation object

```{r}
(cbio <- cBioPortal())
```

## Operations

Check available tags, operations, and descriptions as a `tibble`:

```{r}
tags(cbio)
head(tags(cbio)$operation)
```

## Studies

Get the list of studies available:

```{r}
getStudies(cbio)
```

## Clinical Data

Obtain the clinical data for a particular study:

```{r}
clinicalData(cbio, "acc_tcga")
```

## Molecular Profiles

A table of molecular profiles for a particular study can be obtained by
running the following:

```{r}
mols <- molecularProfiles(cbio, "acc_tcga")
mols[["molecularProfileId"]]
```

## Molecular Profile Data

The data for a molecular profile can be obtained with prior knowledge of
available `entrezGeneIds`:

```{r}
molecularSlice(cbio, molecularProfileId = "acc_tcga_rna_seq_v2_mrna",
    entrezGeneIds = c(1, 2),
    sampleIds = c("TCGA-OR-A5J1-01",  "TCGA-OR-A5J2-01")
)
```

## All available genes

A list of all the genes provided by the API service including hugo symbols,
and entrez gene IDs can be obtained by using the `geneTable` function:

```{r}
geneTable(cbio)
```

It uses the `getAllGenesUsingGET` function from the API.

## Samples

### Sample List Identifiers

To display all available sample list identifiers for a particular study ID,
one can use the `sampleLists` function:

```{r}
sampleLists(cbio, "acc_tcga")
```

### Sample Identifiers

One can obtain the barcodes / identifiers for each sample using a specific
sample list identifier, in this case we want all the copy number alteration
samples:

```{r}
samplesInSampleLists(cbio, "acc_tcga_cna")
```

This returns a `CharacterList` of all identifiers for each sample list
identifier input:

```{r}
samplesInSampleLists(cbio, c("acc_tcga_cna", "acc_tcga_cnaseq"))
```

# sessionInfo

```{r}
sessionInfo()
```

