---
title: "cBioPortalData: User Guide"
author: "Marcel Ramos & Levi Waldron"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{cBioPortalData User Guide}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
bibliography: ../inst/REFERENCES.bib
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

# Installation

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(cBioPortalData)
library(AnVIL)
```

# Introduction

The cBioPortal for Cancer Genomics [website](https://cbioportal.org) is a great
resource for interactive exploration of study datasets. However, it does not
easily allow the analyst to obtain and further analyze the data.

We've developed the `cBioPortalData` package to fill this need to
programmatically access the data resources available on the cBioPortal.

The `cBioPortalData` package provides an R interface for accessing the
cBioPortal study data within the Bioconductor ecosystem.

It downloads study data from the cBioPortal API (the full API specification can
be found here https://cbioportal.org/api) and uses Bioconductor infrastructure
to cache and represent the data.

We demonstrate common use cases of `cBioPortalData` and `curatedTCGAData`
during Bioconductor conference
[workshops](https://waldronlab.io/MultiAssayWorkshop/).

We use the [MultiAssayExperiment][1] (@Ramos2017-er) package to integrate,
represent, and coordinate multiple experiments for the studies available in the
cBioPortal. This package in conjunction with `curatedTCGAData` give access to
a large trove of publicly available bioinformatic data. Please see our
JCO Clinical Cancer Informatics publication [here][2] (@Ramos2020-ya).

[1]: https://dx.doi.org/10.1158/0008-5472.CAN-17-0344
[2]: https://dx.doi.org/10.1200/CCI.19.00119

# Citations

We really appreciate your use and citation of our work. When using
`cBioPortalData`, please cite the following publications:

```{r, include=FALSE}
citation("MultiAssayExperiment")
```

```{r, include=FALSE}
citation("cBioPortalData")
```

You may also want to refer to the `REFERENCES.bib` file in the package:

```{r}
system.file(
    package = "cBioPortalData", "REFERENCES.bib", mustWork = TRUE
)
```

## Overview

### Data Structures

Data are provided as a single `MultiAssayExperiment` per study. The
`MultiAssayExperiment` representation usually contains `SummarizedExperiment`
objects for expression data and `RaggedExperiment` objects for mutation and
CNV-type data.  `RaggedExperiment` is a data class for representing 'ragged'
genomic location data, meaning that the measurements per sample vary.

For more information, please see the `RaggedExperiment` and
`SummarizedExperiment` vignettes.

### Identifying available studies

As we work through the data, there are some datasest that cannot be represented
as `MultiAssayExperiment` objects. This can be due to a number of reasons such
as the way the data is handled, presence of mis-matched identifiers, invalid
data types, etc. To see what datasets are currently not building, we can
look refer to `getStudies()` with the `buildReport = TRUE` argument.

```{r}
cbio <- cBioPortal()
studies <- getStudies(cbio, buildReport = TRUE)
head(studies)
```

The last two columns will show the availability of each `studyId` for
either download method (`pack_build` for `cBioDataPack` and `api_build` for
`cBioPortalData`).

### Choosing download method

There are two main user-facing functions for downloading data from the
cBioPortal API.

* `cBioDataPack` makes use of the tarball distribution of study data. This is
useful when the user wants to download and analyze the entirety of the data as
available from the cBioPortal.org website.

* `cBioPortalData` allows a more flexibile approach to obtaining study data
based on the available parameters such as molecular profile identifiers. This
option is useful for users who have a set of gene symbols or identifiers and
would like to get a smaller subset of the data that correspond to a particular
molecular profile.

## Two main functions

### cBioDataPack: Obtain Study Data as Zipped Tarballs

This function will access the packaged data from \url{cBioPortal.org/datasets}
and return an integrative MultiAssayExperiment representation.

```{r,message=FALSE,warning=FALSE}
## Use ask=FALSE for non-interactive use
laml <- cBioDataPack("laml_tcga", ask = FALSE)
laml
```

### cBioPortalData: Obtain data from the cBioPortal API

This function provides a more flexible and granular way to request a
`MultiAssayExperiment` object from a study ID, molecular profile, gene panel,
sample list.

```{r,warning=FALSE}
acc <- cBioPortalData(api = cbio, by = "hugoGeneSymbol", studyId = "acc_tcga",
    genePanelId = "IMPACT341",
    molecularProfileIds = c("acc_tcga_rppa", "acc_tcga_linear_CNA")
)
acc
```

Note. To avoid overloading the API service, the API was designed to only query
a part of the study data. Therefore, the user is required to enter either a set
of genes of interest or a gene panel identifier.

## Considerations

Note that `cBioPortalData` and `cBioDataPack` obtain data diligently curated
by the cBio Portal data team. The original data and curation lies in the
<https://github.com/cBioPortal/cBioPortal> GitHub repository. However, despite
the curation efforts there may be some inconsistencies in identifiers
in the data. This causes our software to not work as intended though we have
made efforts to represent all the data from both API and tarball formats.

### metadata

You may notice that the `metadata()` may have some additional data that was
not able to be integrated in the `MultiAssayExperiment`.

```{r}
metadata(acc)
```

### Build prompts

You will also get a message for `studyId`s whose data has not been fully
integrated into a `MultiAssayExperiment`.

```{r,echo=FALSE}
cat(
    "Our testing shows that '%s' is not currently building.\n",
    "  Use 'downloadStudy()' to manually obtain the data.\n",
    "  Proceed anyway? [y/n]: y"
)
```

### Manual downloads

For this reason, we have also provided the `downloadStudy`, `untarStudy`, and
`loadStudy` functions to allow researchers to simply download the data and
potentially, manually curate it. Generally, we advise researchers to report
inconsistencies in the data in the [cBioPortal][3] data repository.

[3]: https://github.com/cbioportal/datahub

## Clearing the cache

### cBioDataPack

In cases where a download is interrupted, the user may experience a corrupt
cache. The user can clear the cache for a particular study by using the
`removeCache` function. Note that this function only works for data downloaded
through the `cBioDataPack` function.

```{r,eval=FALSE}
removeCache("laml_tcga")
```

### cBioPortalData

For users who wish to clear the entire `cBioPortalData` cache, it is
recommended that they use:

```{r,eval=FALSE}
unlink("~/.cache/cBioPortalData/")
```

## Example Analysis: Kaplan-Meier Plot

We can use information in the `colData` to draw a K-M plot with a few
variables from the `colData` slot of the `MultiAssayExperiment`. First, we load
the necessary packages:

```{r,message=FALSE,warning=FALSE}
library(survival)
library(survminer)
```

We can check the data to lookout for any issues.

```{r}
table(colData(laml)$OS_STATUS)
class(colData(laml)$OS_MONTHS)
```

Now, we clean the data a bit to ensure that our variables are of the right type
for the subsequent survival model fit.

```{r}
collaml <- colData(laml)
collaml[collaml$OS_MONTHS == "[Not Available]", "OS_MONTHS"] <- NA
collaml$OS_MONTHS <- as.numeric(collaml$OS_MONTHS)
colData(laml) <- collaml
```

We specify a simple survival model using `SEX` as a covariate and we draw
the K-M plot.

```{r}
fit <- survfit(
    Surv(OS_MONTHS, as.numeric(substr(OS_STATUS, 1, 1))) ~ SEX,
    data = colData(laml)
)
ggsurvplot(fit, data = colData(laml), risk.table = TRUE)
```

### Data update requests

If you are interested in a particular study dataset that is not currently
building, please open an issue at our [GitHub repository][4] and we will do our
best to resolve the issues with the code base. Data issues can be opened at
the cBioPortal [data repository][3].
  
[4]: https://github.com/waldronlab/cBioPortalData/issues

We appreciate your feedback!

# sessionInfo

```{r}
sessionInfo()
```

# References
