---
title: "cBioPortalData: API Reference Guide"
author: "Marcel Ramos & Levi Waldron"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{cBioPortal Quick-start Guide}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

# Installation

Please use the devel version of the `AnVIL` Bioconductor package.

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(cBioPortalData)
library(AnVIL)
```

# Introduction

This vignette is for users / developers who would like to learn more about
the available in `cBioPortalData` and possibly hit other endpoints
in the cBioPortal API implementation. The functionality demonstrated
here is used internally by the package to create integrative representations
of study datasets.

# API representation

Obtaining the cBioPortal API representation object

```{r}
(cbio <- cBioPortal())
```

## Operations

Check available tags, operations, and descriptions as a `tibble`:

```{r}
tags(cbio)
head(tags(cbio)$operation)
```

### Searching through the API

```{r}
searchOps(cbio, "clinical")
```

## Studies

Get the list of studies available:

```{r}
getStudies(cbio)
```

## Clinical Data

Obtain the clinical data for a particular study:

```{r}
clinicalData(cbio, "acc_tcga")
```

## Molecular Profiles

A table of molecular profiles for a particular study can be obtained by
running the following:

```{r}
mols <- molecularProfiles(cbio, "acc_tcga")
mols[["molecularProfileId"]]
```

## Molecular Profile Data

The data for a molecular profile can be obtained with prior knowledge of
available `entrezGeneIds`:

```{r}
molecularData(cbio, molecularProfileId = "acc_tcga_rna_seq_v2_mrna",
    entrezGeneIds = c(1, 2),
    sampleIds = c("TCGA-OR-A5J1-01",  "TCGA-OR-A5J2-01")
)
```

## Genes

### All available genes

A list of all the genes provided by the API service including hugo symbols,
and entrez gene IDs can be obtained by using the `geneTable` function:

```{r}
geneTable(cbio)
```

### Gene Panels

```{r}
genePanels(cbio)
getGenePanel(cbio, "IMPACT341")
```

## Molecular Gene Panels

### genePanelMolecular

```{r}
gprppa <- genePanelMolecular(cbio,
    molecularProfileId = "acc_tcga_rppa",
    sampleListId = "acc_tcga_all")
gprppa
```

### getGenePanelMolecular

```{r}
getGenePanelMolecular(cbio,
    molecularProfileIds = c("acc_tcga_rppa", "acc_tcga_gistic"),
    sampleIds = allSamples(cbio, "acc_tcga")$sampleId
)
```

### getDataByGenePanel

```{r}
getDataByGenePanel(cbio, "acc_tcga", genePanelId = "IMPACT341",
    molecularProfileId = "acc_tcga_rppa", sampleListId = "acc_tcga_rppa")
```


It uses the `getAllGenesUsingGET` function from the API.

## Samples

### Sample List Identifiers

To display all available sample list identifiers for a particular study ID,
one can use the `sampleLists` function:

```{r}
sampleLists(cbio, "acc_tcga")
```

### Sample Identifiers

One can obtain the barcodes / identifiers for each sample using a specific
sample list identifier, in this case we want all the copy number alteration
samples:

```{r}
samplesInSampleLists(cbio, "acc_tcga_cna")
```

This returns a `CharacterList` of all identifiers for each sample list
identifier input:

```{r}
samplesInSampleLists(cbio, c("acc_tcga_cna", "acc_tcga_cnaseq"))
```

### All samples within a study ID

```{r}
allSamples(cbio, "acc_tcga")
```

### Info on Samples

```{r}
getSampleInfo(cbio, studyId = "acc_tcga",
    sampleListIds = c("acc_tcga_rppa", "acc_tcga_gistic"))
```

# Advanced Usage

The `cBioPortal` API representation is not limited to the functions
provided in the package. Users who wish to make use of any of the endpoints
provided by the API specification should use the dollar sign `$` function
to access the endpoints.

First the user should see the input for a particular endpoint as detailed
in the API:

```{r}
cbio$getGeneUsingGET
```

Then the user can provide such input:

```{r}
(resp <- cbio$getGeneUsingGET("BRCA1"))
```

which will require the user to 'translate' the response using `httr::content`:

```{r}
httr::content(resp)
```

# sessionInfo

```{r}
sessionInfo()
```

