---
title: "Migrating from cgdsr to cBioPortalData"
author: "Karim Mezhoud & Marcel Ramos"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Migrating from cgdsr to cBioPortalData}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 4
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```


# Introduction

This vignette aims to help developers migrate from the now defunct `cgdsr`
CRAN package. Note that the `cgdsr` package code is shown for comparison but it
is not guaranteed to work. If you have questions regarding the contents,
please create an issue at the [GitHub](https://github.com/waldronlab/cBioPortalData/issues) repository.


# Load the package

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(cBioPortalData)
```

# Discovering studies {.tabset .tabset-fade .tabset-pills}

## `cgdsr`
```{r, comment=FALSE, eval = FALSE}
cgds <- cgdsr::CGDS("http://www.cbioportal.org/")

cgdsr::getCancerStudies.CGDS(cgds)

```

## `cBioPortalData`

* Here we show the default inputs to the `cBioPortal` function for clarity.

```{r , comment=FALSE, warning=FALSE, message=FALSE}
cbio <- cBioPortal(
    hostname = "www.cbioportal.org",
    protocol = "https",
    api. = "/api/api-docs"
)

getStudies(cbio) |> 
   colnames()

```


**Note**

* The `studyId` column is important for further queries.


# Obtaining Cases and Clinical data {.tabset .tabset-fade .tabset-pills}

**Definition**

* The Cases is synonym to the Patients. 

* Each patient is identified  by `Case_id`.

* Each patient has one or multiple samples used for molecular/genetic  profile analysis, like mRNA, mutation, CNA, and Methylation.

## `cgdsr` (Cases)

```{r, eval = FALSE}

cgdsr::getCaseLists.CGDS(x = cgds,
                  cancerStudy = "gbm_tcga_pub")

```
**Note**

* Can not query for multiple  `cancerStudy`

## `cBioPortalData` (Cases)

**Definition**

* Each sample is associated to one or multiple `CaseId`. 

* Each `CaseId` is a Patient. 

* Each patient can participate to one or multiple assays.

* User can find a `case_id` in multiple `case_list_id`.

* The `case_list_description` describes the assays.

* `sampleLists` function uses `studyId` as an argument to return case list identifiers. 

* Be attention to `sampleListId` column.


```{r, echo=TRUE}
sampleLists(api = cbio, 
            studyId = "gbm_tcga_pub") |>
  head(1)
```

* Note that a sample list ID is not required when using the
`fetchAllClinicalDataInStudyUsingPOST` internal endpoint. Data for all
patients can be obtained using the `clinicalData` function.

* It is possible to get directly `case_ids`. We can use `samplesInSampleLists` function.

```{r}
samplesInSampleLists(api = cbio, 
                     sampleListIds = c("gbm_tcga_pub_expr_classical", "gbm_tcga_pub_expr_mesenchymal") )
```

**Note**

* We can query multiple `case_list_id` to get `case_ids` (samples identifiers)

* `sampleInSampleLists` can query a subset of sample list. 

* To get more information about Patients, we can query with `getSampleInfo` function.

```{r}
getSampleInfo(api = cbio, 
              studyId = "gbm_tcga_pub",
              projection = c("SUMMARY", "ID", "DETAILED", "META")) |>
 head(1)
```

There may be other endpoints to get Info. Please reefer to section [Endpoits and Search options][Endpoits and Search options of `cBioportalData`]

```{r}
case_ids_1 <- samplesInSampleLists(api = cbio, 
                                   sampleListIds = "gbm_tcga_pub_expr_classical") |>
                unlist() |>
                head(1) |>
                unname()



cbio$getAllClinicalDataOfSampleInStudyUsingGET(
    sampleId = case_ids_1, 
    studyId = "gbm_tcga_pub"
)


```

## `cgdsr` (Clinic.)

* The User needs to specify which samples list from one Study.

* `getClinicalData`  uses `case_list_id` as an argument without specifying the `study_id`.

```{r, eval = FALSE}
getClinicalData.CGDS(x = cgds, 
                     caseList =  "gbm_tcga_pub_expr_classical")

```

**Note**

* We queried clinical data for `_expr_classical` cases in `gbm_tcga_pub` study id.

## `cBioPortalData` (Clinic.)

* Compared to `cgdsr`, `clinicalData` function needs `studyId` and not `case_list_id` to query clinical data sets.

```{r}
## get Clinical data
clinicalData(api = cbio, 
             studyId = "gbm_tcga_pub") |>
  colnames()
```

**Note**

* Explicitly we queried a subset (only `_expr_classical`, 54 samples) clinical data with `cgdsr` and all (206 samples) clinical data with `cBioPortalData`.

* `cgdsr` returns a dataframe with `sampleId` (TCGA.02.0009.01)  but not `patientId` (TCGA.02.0009)

* `cBioPortalData` returns `sampleId` (TCGA-02-0009-01) & `patientId` (TCGA-02-0009).


# get Genetic Profiles or Molecular Profiles Identifiers {.tabset .tabset-fade .tabset-pills}

## `cgdsr`
```{r, eval=FALSE}
getGeneticProfiles.CGDS(cgds,cancerStudy = "gbm_tcga_pub" )
```

## `cBioPrtalData`

```{r}
molecularProfiles(api = cbio, studyId = "gbm_tcga_pub") |>
  head(1)
```

# Profile Data or molecular Data (mRNA expression) for a set of  genes {.tabset .tabset-fade .tabset-pills}

## `cgdsr`

```{r, eval=FALSE}
getProfileData.CGDS(x = cgds,
                    genes = c("NF1", "TP53", "ABL1"),
                    geneticProfiles = "gbm_tcga_pub_mrna",
                    caseList = "gbm_tcga_pub_all") %>%
                     datatable(options = list(pageLength = 5))
```

## `cBioPortalData`

Currently, some conversion is needed to directly use the `molecularData`
function, if you only have Hugo symbols. First, convert to Entrez gene IDs
and then obtain all the samples in the sample list of interest.

### Convert gene HugoSymbol to Entrez

```{r, eval=TRUE}
## convert Hugo Symbol genes to Entrez code
our_entrez <- queryGeneTable(cbio,
                             by = "hugoGeneSymbol",
                             genes = c("NF1", "TP53", "ABL1")
                             )%>%
                             pull(entrezGeneId)
out_entrez
```

**Convert all or selected `genePanelId` to gene Hugo Symbol:**

```{r, eval = FALSE}
# get all genPanelId
all_genePanelId <- genePanels(api = cbio) %>% pull(genePanelId)


## get all Genes entrez/symbol from all genePanelID, rm duplicates
lapply(X = all_genePanelId,
       function(x) getGenePanel(api = cbio, 
                                genePanelId = x)) %>%
                 bind_rows() %>%
                 distinct() %>%
    datatable(options = list(pageLength = 5))
```

### Three ways to get molecular Data for mRNA expression

**(1) `molecularData` function**
```{r}

## get mRNA expression
molecularData(api = cbio, 
              molecularProfileIds = "gbm_tcga_pub_mrna",
              entrezGeneIds = c("7157", "25", "4763"), # our_entrez, 
              sampleIds = unlist(samplesInSampleLists(api = cbio,
                                                      sampleListIds = "gbm_tcga_pub_all"))
              )
```

**(2) `getDataByGenes` function**

* Note that this can also be done by using the `getDataByGenes` function, which
automatically figures out all the sample identifiers in the study:

```{r}
our_mRNA_gbm <- getDataByGenes(
    api =  cbio,
    studyId = "gbm_tcga_pub",
    genes = c("NF1", "TP53", "ABL1"),
    by = "hugoGeneSymbol",
    molecularProfileIds = "gbm_tcga_pub_mrna"
)

assay(our_mRNA_gbm[["gbm_tcga_pub_mrna"]])[, 1:4]
```

**(3) `cBioPortalData` function**
It is important to note that end users who wish to obtain the data as
easily as possible should use the main `cBioPortalData` function:

```{r, eval=TRUE}
our_gbm_pub <- cBioPortalData(
    api = cbio,
    studyId = "gbm_tcga_pub",
    genes = c("NF1", "TP53", "ABL1"), 
    by = "hugoGeneSymbol",
    molecularProfileIds = "gbm_tcga_pub_mrna"
)
assay(our_gbm_pub[["gbm_tcga_pub_mrna"]])[, 1:4]
```

# Mutation Data {.tabset .tabset-fade .tabset-pills}

## `cgdsr`

```{r, eval = FALSE}
getMutationData.CGDS(
    x = cgds,
    caseList = "getMutationData",
    geneticProfile = "gbm_tcga_pub_mutations",
    genes = c("NF1", "TP53", "ABL1")
)
```


## `cBioPortalData`

### Three ways to get molecular Data for mutation

**`mutationData` function**

Similar to `molecularData`, `mutationData` function  can query mutation data.
```{r, eval=TRUE}
our_Mut_gbm <- mutationData(
    api = cbio,
    molecularProfileIds = "gbm_tcga_pub_mutations",
    by = "hugoGeneSymbol",
    genes = c("NF1", "TP53", "ABL1"),
    #entrezGeneIds = c("7157", "25", "4763"), # our_entrez, 
    sampleIds = unlist(samplesInSampleLists(api = cbio,
                                                      sampleListIds = "gbm_tcga_pub_all"))
) 

assay(our_Mut_gbm[["gbm_tcga_pub_mutations"]])[, 1:4]
```

**(2) `getDataByGene` function**

 Mutation data can also queried by  the `getDataByGenes` function.

```{r}
our_Mut_gbm <- getDataByGenes(
    api = cbio,
    studyId = "gbm_tcga_pub",
    genes = c("NF1", "TP53", "ABL1"),
    by = "hugoGeneSymbol",
    molecularProfileIds = "gbm_tcga_pub_mutations"
) 

assay(our_Mut_gbm[["gbm_tcga_pub_mutations"]])[, 1:4]
```

**(3) `cBioPortalData` function**

```{r, eval=TRUE}
our_Mut_gbm <- cBioPortalData(
    api = cbio,
    studyId = "gbm_tcga_pub",
    genes = c("NF1", "TP53", "ABL1"),
    by = "hugoGeneSymbol",
    molecularProfileIds = "gbm_tcga_pub_mutations"
)

assay(our_Mut_gbm[["gbm_tcga_pub_mutations"]])[, 1:4]
```

# Copy Number Alteration Data (CNA)  {.tabset .tabset-fade .tabset-pills}

## `cgdsr`

## `cBioPortalData`

# Methylation Data  {.tabset .tabset-fade .tabset-pills}

## `cgdsr`

## `cBioPortalData`

# Endpoits and Search options of `cBioportalData`

* `searchOps` function returns all endpoints needed using `regex`. Here some example of endpoints:  


**By Sample endpoint**
```{r}
searchOps(cbio, "sample")
```



**By Clinical endpoint**
```{r}
searchOps(cbio, "clinical")
```


**By molecular endpoint**
```{r}
searchOps(cbio, "molecular")
```

**By profile endpoint**
```{r}
searchOps(cbio, "profile")
```

**By mutation endpoint**

```{r}
searchOps(cbio, "mutation")
```
